Neural Connectome of the Ctenophore Statocyst

### 2025-07-09 manuscript/Manuscript.qmd
MS: edits of intro

A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.

>>>>

A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy (vEM) and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.
